CF Nutrition Pre- and Post-CFTR Modulator

### 2022-09-29 Edith Zemanick/CF Nutrition/cf_nutrition.qmd
Update cf_nutrition.qmd

```{r}
# Import all data
df = read.csv("/Users/timvigers/Documents/Work/Vigers/CF/Edith Zemanick/CF Nutrition/Data_Cleaned/analysis.csv")
df$Timepoint = factor(df$Timepoint,levels = c("Pre","Post"))
```

@@

-   We excluded visits more than 3 years prior to effective modulator start and within 90 after effective modulator start.

-   All values prior to effective modulator start were averaged to calculate the "pre" value. We selected the most recent vitamin values as the "post" value, and included any BMI percentile, FEV1 % predicted, ALT, AST, and GGTP measures within a month of that visit.

@@

@@

```{r}
t1 = tableby(First.Modulator~Vitamin.A+Vitamin.D+Vitamin.E.Alpha+
           ppFEV1+BMI.Percentile+ALT+AST+GGTP,data = df[df$Timepoint == "Pre",])
summary(t1,pfootnote = T)
```

@@

```{r results='asis'}
p = paired(Timepoint~Vitamin.A+Vitamin.D+Vitamin.E.Alpha+
           ppFEV1+BMI.Percentile+ALT+AST+GGTP,data = df,id=ID,control = list(numeric.test = "signed.rank"))
summary(p,pfootnote = T)
```

>>>>

```{r}
# Import datasets
vit <- read_excel("/Users/timvigers/Documents/Work/Vigers/CF/Edith Zemanick/CF Nutrition/Data_Raw/Report Vitamins_throughDecember2021_revised20220712.xlsx")
mods <- read_excel("/Users/timvigers/Documents/Work/Vigers/CF/Edith Zemanick/CF Nutrition/Data_Raw/Report CFTR Modulators_throughDecember2021.xlsx")
pfts <- read_excel("/Users/timvigers/Documents/Work/Vigers/CF/Edith Zemanick/CF Nutrition/Data_Raw/Report_PFTs_ThroughDecember2021.xlsx")
lfts <- read_excel("/Users/timvigers/Documents/Work/Vigers/CF/Edith Zemanick/CF Nutrition/Data_Raw/Report_LFTs_ThroughDecember2021.xlsx")
pi = read_excel("/Users/timvigers/Documents/Work/Vigers/CF/Edith Zemanick/CF Nutrition/Data_Raw/Report Fecal Elastase_throughDecember2021.xlsx")
# Format
vit$VitaminA_Retinol = as.numeric(vit$VitaminA_Retinol)
vit$`Age at Test (years)` <- NULL
vit$Date = as.Date(vit$Date)
vit$`Vitamin E Gamma` = NULL
mods$`Earliest Date in Registry` = as.Date(mods$`Earliest Date in Registry`)
pfts <- pfts %>%
  select(PatientID, Sex, Date, `Height (cm)`:`FEV1 % pred`) %>%
  rename(`Patient ID` = PatientID)
pfts$Date = as.Date(pfts$Date)
lfts <- lfts %>% select(`Patient ID`, Date:GGTP)
lfts$Date = as.Date(lfts$Date)
pi = pi %>% select(`Patient ID`,Date,`Fecal Elastase`)
pi$`Fecal Elastase` = as.numeric(pi$`Fecal Elastase`)
pi$Date = as.Date(pi$Date)
# Get first modulator and start date
mods$`Earliest Date in Registry` <- parse_date(mods$`Earliest Date in Registry`, approx = F)
mods <- mods %>%
  filter(Modulator %in% c("Kalydeco", "Trikafta")) %>%
  pivot_wider(
    id_cols = `Patient ID`, names_from = Modulator,
    values_from = c(`Earliest Date in Registry`,`Age at Start (years)`)
  )
mods$Start <- pmin(mods$`Earliest Date in Registry_Trikafta`, 
                   mods$`Earliest Date in Registry_Kalydeco`, na.rm = T)
mods$tf <- mods$`Earliest Date in Registry_Trikafta` == mods$Start
mods$kf <- mods$`Earliest Date in Registry_Kalydeco` == mods$Start
mods <- mods %>% unite(First, tf, kf, na.rm = T)
mods$`Age at Start` = pmin(mods$`Age at Start (years)_Trikafta`,mods$`Age at Start (years)_Kalydeco`,na.rm = T)
mods$First <- factor(mods$First,
  levels = c("FALSE_TRUE", "TRUE"),
  labels = c("Kalydeco", "Trikafta")
)
# Combine
vit = full_join(vit,pfts, by = c("Patient ID", "Date"))
vit = full_join(vit,lfts, by = c("Patient ID", "Date"))
vit = full_join(vit,pi,by = c("Patient ID", "Date"))
vit <- left_join(vit, mods %>% select(`Patient ID`, Start, First,`Age at Start`), 
                 by = "Patient ID")
# Get pre vs. post info, time from start
vit$`Pre-/Post-` <- factor(vit$Date >= vit$Start,
  levels = c("FALSE", "TRUE"),
  labels = c("Pre", "Post")
)
vit$`Days from Start` = as.numeric(difftime(vit$Date,vit$Start,units = "days"))
# Filter those more than 3 years before start, and within 90 days after start
vit = vit %>% filter(`Days from Start` > -3*365,
                     `Days from Start` < 0 | `Days from Start` >= 90)
# Order variables, fill sex
vit = vit %>% select(`Patient ID`,Sex,Date,Start,First,`Age at Start`,
                     `Pre-/Post-`,`Days from Start`,everything()) %>% 
  group_by(`Patient ID`) %>% 
  fill(Sex,.direction = "downup")
# Average pre-mod values, get most recent vitamin values and and any other variables within 30 days
pre = vit %>% group_by(`Patient ID`) %>% 
  filter(`Pre-/Post-` == "Pre",
         !is.na(VitaminA_Retinol) | !is.na(`25OH-Vitamin D`) | !is.na(`Vitamin E Alpha`)) %>%
  mutate(n=n()) %>% filter(n>=2) %>% # At least two pre measures
  summarise(`First Modulator` = as.character(First[1]),Timepoint = "Pre",
            `Age at Start` = `Age at Start`[1],Sex = Sex[1],
            across(`Days from Start`:`Fecal Elastase`, ~ mean(.x, na.rm = TRUE)))
post = vit %>% group_by(`Patient ID`) %>% 
  filter(`Pre-/Post-` == "Post",
         !is.na(VitaminA_Retinol) | !is.na(`25OH-Vitamin D`) | !is.na(`Vitamin E Alpha`)) %>%
  mutate(latest = max(Date),date_cutoff= latest - lubridate::days(30)) %>%
  filter(Date >= date_cutoff) %>% 
  arrange(`Patient ID`,Date) %>%
  summarise(`First Modulator` = as.character(First[1]),Timepoint = "Post",
            `Age at Start` = `Age at Start`[1],Sex = Sex[1],
            across(`Days from Start`:`Fecal Elastase`,
                   ~if(all(is.na(.))) NA else last(na.omit(.))))
df = bind_rows(pre,post) %>%
  mutate_all(~ifelse(is.nan(.), NA, .))
df$Timepoint = factor(df$Timepoint,levels = c("Pre","Post"))
# Pancreatic status and LFTs
df$`Pancreatic Insufficient` = factor(df$`Fecal Elastase` < 128,
                       levels = c("TRUE","FALSE"),labels = c("Yes","No"))
df$`Elevated LFTs (3X)` = factor((!is.na(df$ALT) & df$ALT > 3*55) | 
                                   !is.na(df$AST) & df$AST > 3*48 | 
                                   !is.na(df$GGTP) & df$GGTP > 3*61,
                                 levels = c("TRUE","FALSE"),labels = c("Yes","No"))
df$`Elevated LFTs (5X)` = factor((!is.na(df$ALT) & df$ALT > 5*55) | 
                                   !is.na(df$AST) & df$AST > 5*48 | 
                                   !is.na(df$GGTP) & df$GGTP > 5*61,
                                 levels = c("TRUE","FALSE"),labels = c("Yes","No"))
```

@@

-   We excluded visits more than 3 years prior to effective modulator start and within 90 days after effective modulator start.

-   All values prior to effective modulator start were averaged to calculate the "pre" value. We selected the most recent vitamin values as the "post" value, and included any BMI percentile, FEV1 % predicted, ALT, AST, and GGTP measures within 30 days of that visit.

@@

-   Pancreatic insufficiency was determined based on a fecal elastase cutoff of \< 128 $\mu g/g$.

-   Elevated liver labs were considered at two levels:

    -   ALT, AST, or GGTP 3 times the normal level

    -   ALT, AST, or GGTP 5 times the normal level

-   Normal liver labs were considered:

    -   ALT: 55 U/L

    -   AST: 48 U/L

    -   GGTP: 61 U/L

@@

```{r results='asis'}
t1 <- tableby(`First Modulator` ~ `Age at Start` + Sex + `VitaminA_Retinol` + `25OH-Vitamin D` + `Vitamin E Alpha` + `FEV1 % pred` + `BMI percentile (CDC)` + `ALT` + `AST` + `GGTP` + `Elevated LFTs (3X)` + `Elevated LFTs (5X)` + `Fecal Elastase` + `Pancreatic Insufficient`,data = df[df$Timepoint == "Pre", ],numeric.stats = c("Nmiss","medianrange"))
summary(t1, pfootnote = T)
```

@@

```{r results='asis'}
p <- paired(Timepoint ~ `VitaminA_Retinol` + signed.rank(`25OH-Vitamin D`,"Nmiss","median","q1q3") + `Vitamin E Alpha` + `FEV1 % pred` + `BMI percentile (CDC)` + `ALT` + `AST` + `GGTP` + `Elevated LFTs (3X)` + `Elevated LFTs (5X)` + `Pancreatic Insufficient`,data = df, id = `Patient ID`)
summary(p, pfootnote = T)
```
